Prefetching SPARQL query cacher

### 2016-07-28 dissertation/cache-proxy.tex
describe join rotation

- NestedLoop Join (distinct cost: 180)
-   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
-   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
\end{verbatim}
\end{example}
In the above example, the first plan tree will be returned from the
query planner for execution, since it has the lowest estimated cost
(27).

@@

Following this discussion, consider this example:
\begin{example}{A four-triple pattern BGP}
\begin{verbatim}
?s <http://example.org/m/p> "1" ;
   <http://example.org/m/q> <http://example.org/m/a> ;
   <http://example.org/m/p> ?o .
?o <http://example.org/m/b> "2" .
\end{verbatim}
\end{example}
Say that the first triple pattern has a result in the cache. This will
not cause Problem~\ref{prob:cachecartesian} to occur (it would have
occurred if the third triple pattern had a cached result, however), so
we can safely use the cache. The results of the three remaining triple
patterns must be fetched from a remote server. There are essentially
two reasonable choices: Either fetch the result from an LDF server as
argued by \cite{verborgh2014querying} and join them locally or send a
single, rewritten SPARQL query to the remote endpoint for the
evaluation. When the results are returned from both the cache and the
remote endpoint, they must then finally be joined to produce the final
result returned to the client. A string representation of the plan
tree from the current implementation reflects this and reads (note
that accessing the cache is represented by an iterator plan):
\begin{example}{An example Attean plan with a remotely executed BGP}
\begin{verbatim}
- Hash Join { s } (distinct cost: 74)
-   SPARQLBGP (cost: 60)
-     Quad { ?s, <http://example.org/m/p>, ?o, <http://test.invalid/graph> }
-     Quad { ?o, <http://example.org/m/b>, "2", <http://test.invalid/graph> }
-     Quad { ?s, <http://example.org/m/q>, <http://example.org/m/a>, <http://test.invalid/graph> }
-   Iterator (?s with 2 elements) (cost: 2)
\end{verbatim}
\end{example}

>>>>

- NestedLoop Join (distinct cost: 180)
-   Quad { ?s, <http://example.org/p>, ?o, <http://example.org/> } (cost: 15)
-   Quad { ?s, <http://example.org/p>, "1", <http://example.org/> } (cost: 12)
\end{verbatim}
\end{example}
In the above example, the first plan tree will be returned from the
query planner for execution, since it has the lowest estimated cost
(27). Since the SPARQL engine operates over quads, we have assigned a
dummy graph name when constructing the Model, namely
\rdfterm{<http://example.org/>}.

@@

Following this discussion, consider this example:
\begin{example}{A four-triple pattern BGP}\label{ex:bgp4}
\begin{verbatim}
?s <http://example.org/m/p> "1" ;
   <http://example.org/m/q> <http://example.org/m/a> ;
   <http://example.org/m/p> ?o .
?o <http://example.org/m/b> "2" .
\end{verbatim}
\end{example}
Say that the first triple pattern has a result in the cache. This will
not cause Problem~\ref{prob:cachecartesian} to occur (it would have
occurred if the third triple pattern had a cached result, however), so
we can safely use the cache. The results of the three remaining triple
patterns must be fetched from a remote server. There are essentially
two reasonable choices: Either fetch the result from an LDF server as
argued by \cite{verborgh2014querying} and join them locally or send a
single, rewritten SPARQL query to the remote endpoint for the
evaluation. When the results are returned from both the cache and the
remote endpoint, they must then finally be joined to produce the final
result returned to the client. A string representation of the plan
tree from the current implementation reflects this and reads (note
that accessing the cache is represented by an iterator plan):
\begin{example}{An example Attean plan with a remotely executed BGP}
\begin{verbatim}
- Hash Join { s } (distinct cost: 74)
-   SPARQLBGP (cost: 60)
-     Quad { ?s, <http://example.org/m/p>, ?o, <http://test.invalid/graph> }
-     Quad { ?o, <http://example.org/m/b>, "2", <http://test.invalid/graph> }
-     Quad { ?s, <http://example.org/m/q>, <http://example.org/m/a>, <http://test.invalid/graph> }
-   Iterator (?s with 2 elements) (cost: 2)
\end{verbatim}
\end{example}

Merging the quad plans into a single SPARQL BGP plan is not
straightforward in the presence of a quad pattern that is to be looked
up in e.g. a cache. If we assign the letters $A$, $B$, $C$ and $D$ to
quad patterns\todo{correctly expressed?}  in the order in
Example~\ref{ex:bgp4}, the initial join tree produced by parsing may
be $ ((A \bowtie B )\bowtie C) \bowtie D) $. Now, we are interested in
a plan on the form $A \bowtie BCD$ (all commutations will be
generated). To facilitate this, we employ \emph{join rotation}. A
simple example of join rotation would be this: We first generate
subplans for both $((A \bowtie B )\bowtie C) \bowtie D) $ and $ (A
\bowtie B )\bowtie (C \bowtie D) $. Then, $C$ and $D$ can be merged to
a SPARQL BGP to $ (A \bowtie B )\bowtie CD $. Another rotation can
then be performed: $ A \bowtie (B \bowtie CD) $ and then coalesced to
the final result. The planner will generate alternative plans for both
rotated and unrotated plans, and cost is estimated for both, so we
have to be careful to design the cost model to allow the rotation to
result in SPARQL BGP plans.
